---
title: "Week 09 LOESS"
author: "Xavier Francis"
date: "2025-03-10"
output: html_document
---

What do you do if your data are nonlinear?

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(lattice)
library(broom) #make nice summary outputs
library(rvest) #webscraping
library(lubridate) #handle dates/times
#install.packages("ggfortify")
library(ggfortify)

```
# Plot example data "ethanol"
https://search.r-project.org/CRAN/refmans/SemiPar/html/ethanol.html
"The ethanol data frame contains 88 sets of measurements for variables from an experiment in which ethanol was burned in a single cylinder automobile test engine."

NOx=nitrogen oxide
E=Ethanol

```{r}
head(ethanol)

#plot just the points
ggplot(ethanol, aes(E, NOx)) +
  geom_point()

```
#add regression line to the plot

```{r}
ggplot(ethanol, aes(E, NOx)) +
  geom_point() + 
  geom_smooth(method='lm')



# run a regression
lmfit <- lm(NOx ~ E, ethanol)
summary(lmfit)
```

## Use broom package to view a nicer layout of the regression results
```{r}
broom::tidy(lmfit)
broom::glance(lmfit)

```
## Check diagnostics to assess linearity

```{r}
autoplot(lmfit)

```
From: https://www.sthda.com/english/articles/39-regression-model-diagnostics/161-linear-regression-assumptions-and-diagnostics-in-r-essentials/

"The diagnostic plots show residuals in four different ways:

Residuals vs Fitted. Used to check the linear relationship assumptions. A horizontal line, without distinct patterns is an indication for a linear relationship, what is good.

Normal Q-Q. Used to examine whether the residuals are normally distributed. It’s good if residuals points follow the straight dashed line.

Scale-Location (or Spread-Location). Used to check the homogeneity of variance of the residuals (homoscedasticity). Horizontal line with equally spread points is a good indication of homoscedasticity. This is not the case in our example, where we have a heteroscedasticity problem.

Residuals vs Leverage. Used to identify influential cases, that is extreme values that might influence the regression results when included or excluded from the analysis. This plot will be described further in the next sections."


CONCLUSION: NONLINEAR



# LOESS Regression

A LOESS/LOWESS (Locally Weighted Scatter-plot Smoother) regression involves fitting a smooth curve between two or more points in a series.


## Animation of fitting loess curves

http://varianceexplained.org/files/loess.html



# LOESS example
```{r}
#set the ggplot theme to black and white
theme_set(theme_bw())

#do a loess fit
mod <- loess(NOx ~ E, ethanol, degree = 1, span = 0.5)
summary(mod)

#add information to the model with augment
fit <- broom::augment(mod)
fit

ggplot(fit, aes(E, NOx)) +
  geom_point() +
  geom_line(aes(y = .fitted), color = "red")


```
# Weather data example

## Scrape yesterday's weather data
```{r}
col_link <- "https://www.wunderground.com/dashboard/pws/KTXDRIPP119/table/2025-03-09/2025-03-09/daily"
col_page <- read_html(col_link)

ds.weather <- col_page %>% html_nodes("table.history-table.desktop-table") %>%
  html_table() %>%.[[1]]


#make temperature numeric
ds.weather$temp <- as.numeric(substr(ds.weather$Temperature,1,nchar(ds.weather$Temperature)-3))


ggplot(ds.weather, aes(x=Time, y=temp))+
         geom_point()


```

# Fix time formatting
```{r}

#extract just the morning data
ds.weatherAM <- ds.weather[grepl("AM", ds.weather$Time), , drop = FALSE]


#make a decimal version of the time
t_hm <- lubridate::hm(ds.weatherAM$Time)
t_hm_dec <- lubridate::hour(t_hm) + lubridate::minute(t_hm)/60

ds.weatherAM$dec.time <- t_hm_dec

#view decimal time vs original time
ds.weatherAM[,c(1,14)]
```

## Plot
```{r}
ggplot(ds.weatherAM, aes(x=dec.time, y=temp))+
         geom_point()
```
## fix 12AM issue
```{r}
ds.weatherAM$dec.time <- ifelse(ds.weatherAM$dec.time>=12, ds.weatherAM$dec.time-12, ds.weatherAM$dec.time)
```



## Plot temp data with decimal time, 12AM issue fixed
```{r}
ggplot(ds.weatherAM, aes(x=dec.time, y=temp))+
         geom_point()+
         geom_line()
```

## Add linear model to plot
```{r}
# run a regression
lmfit <- lm(temp ~ dec.time, ds.weatherAM) #formula y~x, data
summary(lmfit)

# Use broom package to view a nicer layout of the regression results
summary <- broom::glance(lmfit)
format(summary$p.value,scientific = FALSE)
round(summary$p.value, 4)

#use ifelse to get nice p-value for plot title
pval <- ifelse(round(summary$p.value, 4)>0, round(summary$p.value, 4), "<0.0001")

ggplot(ds.weatherAM, aes(x=dec.time, y=temp))+
         geom_point()+
         geom_line() +
        geom_smooth(method='lm') +
        labs(title=paste0("Linear regression R2=", round(summary$r.squared,2), " (p-value", pval, ")"))
```

# Fit a smoothed line with LOESS
```{r}
#do a loess fit
mod <- loess(temp ~ dec.time, ds.weatherAM, degree = 1, span = 0.75)
summary(mod)

#add information to the model with augment
fit <- broom::augment(mod)
fit

ggplot(fit, aes(dec.time, temp)) +
  geom_point() +
  geom_line() +
  geom_smooth(method='lm') +
  geom_line(aes(y = .fitted), color = "red")+
  labs(title="Black=observed data, Blue=linear regression fit, Red=LOESS fit (span 0.75)")
```



# Loop to find 5 spans

Plan: Create vector of spans
Create empty df that we can store results into
within the loop:
do the loess fit for the particualar span
add the extra information to the model with augment
add a new column with a span ID
rbind the data frame onto the master dataframe
end loop
convert span ID into a character for plotting purposes
plot spans

```{r}
# KEY
#specify vectors of spans
spans <- c(.9,.75,.6,.5,.4)


# make blank dataframe to write loop output into
span_df <- data.frame()


# LOOP
#do a loess fit
  for (i in 1:length(spans)){

#loess fit     
mod<- loess(temp ~ dec.time, ds.weatherAM, degree = 1, span = spans[i])

#add information to the model with augment
fit <- broom::augment(mod)
#add column of spans to differentiate data for each span
fit$spanID <- spans[i]
#bind span specific loop dataframe to the master dataframe
span_df <- rbind(span_df, fit)
    
  }
    
#convert span to a character for plotting
span_df$spanID <- as.character(span_df$spanID ) 


# Show written data
glimpse(span_df)
```


#plot the data with the various spans
```{r}

#plot the data
 ggplot(span_df, aes(dec.time, temp, group = spanID)) +
  geom_point() +
  geom_line() +
  geom_smooth(method='lm') +
  geom_line(aes(y = .fitted, group = spanID, colour = spanID), linewidth = 1)+
  labs(title="Black=observed data, Blue=linear regression fit, Multicolors=spans")
   
```




#span .4 seems to fit the original data the closest


