---
title: "ADA hw 3"
author: "Xavier Francis"
date: "2025-01-30"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```


# Generate data
## random dates
```{r}
set.seed(42)
date <- sample(seq(as.Date('2023-01-01'), as.Date('2025-01-20'), by = "day"), 400)

```

## Random measures for two separate groups
Assume we have some knowledge of the typical sizes of individuals of both groups. We can generate random measures around a mean value with a set amount of variability (standard )
```{r}
# Set the seed for reproducibility
set.seed(42)
 
# Generate 1000 random numbers from a normal distribution
mean_value <- 23
sd_value <- 2
num_samples <- 400
random_numbers1 <- rnorm(num_samples, mean = mean_value, sd = sd_value)

set.seed(42)
mean_value2 <- 37
sd_value2 <- 4
#num_samples <- 400
random_numbers2 <- rnorm(num_samples, mean = mean_value2, sd = sd_value2)


```

## Combine and randomize the two groups
```{r}
#combine the two groups
measure <- c(random_numbers1, random_numbers2)


#randomize the order of the values in the measure vector
set.seed(42) # just to make it reproducible
measure <- sample(measure)

```

## Generate random colors
```{r}
color <- rep(c("orange", "yellow"), each=200)


#randomize the values in the color vector
set.seed(42) # just to make it reproducible
color <- sample(color)

```


## combine all data into single dataframe & write data
```{r}
#combine data into dataset
data <- cbind.data.frame(date, measure, color)

#write data 


#write.csv(data, "C:/Users/Xmanf/Desktop/Advanced data analysis stuff/W03.randomdata_X.csv")



```

# Mess it up then reimport
Outside of R, intentionally introduce mistakes or messiness into the data. Then reimport it here
```{r}
#import messed up data
messy <- read_csv("W03.randomdata_messy.csv")

```
#possible sourcesm of error in the dataset
color date FoRMAT outliers whitespace

```{r}
library(readxl)
#rename data
messy <- w03_randomdata_Roya_messy
#find unique values in color column so we can find what values to fix
unique(messy$color)
#color fix. replace mispelling with correct color
messy$color[messy$color == "oragne"] <- "orange"
unique(messy$color)

#take a look at date and measure to spot check for messed up/missing values
messy$date
(messy$measure)
 
#check for NAs in measure
is.na(messy$measure)

#plot a boxplot to visualize data
ggplot(data =messy, aes(y = measure)) + geom_boxplot()

#restrict data within a reasonable range to get rid of obvious errors in the measure data points 
reasonable <- subset(messy, messy$measure < 5000)

#remove nas
reasonable_no_na <- na.omit(reasonable)

#reasonable_no_na <- reasonable[!(is.na(reasonable$measure)), ]


library(lubridate)

#convert character to date format
reasonable_no_na$date.fixed <- as.Date(reasonable_no_na$date, format="%m/%d/%Y")


```
#Statistically compare measure for orange vs yellow
```{r}
head(data)

ggplot(data =data, aes(x = color, y = measure)) + geom_boxplot()

#t test
t.test(measure ~ color, data = data)

#anova
oneway.test(measure ~ color, data = data, var.equal = TRUE)
#p-value = 0.8777



```


