---
title: "ADA_HW_week4"
author: "Xavier Francis"
date: "2025-02-07"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Load packages and import data with unknown true clusters
```{r}
library(tidyverse)

kmean_thing <- read.csv("C:/Users/Xmanf/Desktop/Advanced data analysis stuff/w04.unknown.clusters.csv", header=TRUE, stringsAsFactors=FALSE)

```

#get a sense of the structure of the dataset
```{r}

glimpse(kmean_thing)
```


#visualization of raw data for initial investigation 
```{r}

#plot raw data using original clusters  
ggplot(kmean_thing, aes(x = as.factor(cluster), y = height, fill = as.factor(cluster))) +
  geom_violin(alpha = 0.5, show.legend = FALSE, fill = "gray") +
  geom_boxplot(width = 0.05) +
  theme(legend.position = "none") +
  labs(x = "cluster", y = "height") +
  theme_minimal() 
  #scale_y_continuous(breaks = seq(0, 100, 50) +  
     #coord_cartesian(ylim = c(0, 50))

  #plot raw data without any clusters
  ggplot(kmean_thing, aes(x = "", y = height)) +
  geom_violin(alpha = 0.5, show.legend = FALSE, fill = "gray") +
  geom_boxplot(width = 0.05) +
  theme(legend.position = "none") +
  labs(x = "cluster", y = "height") +
  theme_minimal() 
  
  
```
#visualize data via a histogram to investigate possible clusters
```{r}
#Alternate to try and visualize possible clusters
  ggplot(kmean_thing, aes(x = height)) +
geom_histogram(color = "black", fill = "white", bins = 150)  
  
```
#plot dendrogram
```{r} 
#Alternate to try and visualize possible clusters
#install.packages("ggdendro")
library(ggdendro)
hc <- hclust(dist(kmean_thing$height), "ave")
ggdendrogram(hc, rotate = TRUE, size =1)
```

#add cluster id and plot again
```{r}
#rename cluster column to initial cluster
colnames(kmean_thing)[2] <- 'initial_cluster'

#cluster data using kmeans
kmean_thing.wclust <- cbind(kmean_thing, cluster = kmeans(kmean_thing$height, 5)$cluster)

#visualize data as a violin plot with new clusters
ggplot(kmean_thing.wclust, aes(x = as.factor(cluster), y =height, fill = as.factor(cluster))) +
  geom_violin(trim = FALSE, fill = "gray") +
  geom_boxplot(width = 0.05) +
  theme(panel.border = element_rect(colour = "black", fill = NA, linewidth = 0.5))

```


#kmeans parts
```{r}
names(kmeans(kmean_thing$height,5))
kmeans(kmean_thing$height, 5)$size #how many values at each clusters
kmeans(kmean_thing$height, 5)$centers # what is the center of each clsuter?
```




#create key for loop
```{r}
#create key for loop for creating randomized data
KEY.loop <- c("Izzys", "Birthday", 0,6,2,4,1,9,9,7 )

for(KEY in 1:length(KEY.loop)){#KEY = 1
  loop.id <- KEY.loop[KEY]
  print(loop.id)}
  
```
#generate clusters of randomized data with particular mean and particular sd using loop
```{r}
KEY_means <- c(45.5,66,24,10,300,200,124,500)
#set random parameters so theyre the same for each run
sd_value <- 1
num_samples <- 200
master_random_numbers <- data.frame()
#master_random_numbers <- c()
for(k in 1:length(KEY_means)){ #k = 1
  loop.mean <- KEY_means[k]
  #set seed
  set.seed(11)
random_numbers <- rnorm(num_samples, mean = loop.mean, sd = sd_value)
random_numbers <- as.data.frame(random_numbers)
master_random_numbers <- rbind.data.frame(master_random_numbers, random_numbers)
  #master_random_numbers <- c(master_random_numbers, random_numbers )
}



```

#visualize randomized data
```{r}

ggplot(master_random_numbers, aes( y =random_numbers, x = "")) +
  geom_violin(trim = FALSE, fill = "gray") +
  geom_boxplot(width = 0.05) +
  theme(panel.border = element_rect(colour = "black", fill = NA, linewidth = 0.5))

```


#Kmeans calculation
```{r}

#KEY_means <- c(45.5,66,24,10,300,200,124,500)


#obtain clusters from kmeans
master_random_numbers_kmean <- cbind(master_random_numbers, cluster = kmeans(master_random_numbers$random_numbers, 8)$cluster)

#visualize data with kmeans clusters
ggplot(master_random_numbers_kmean, aes(x = as.factor(cluster), y =random_numbers, fill = as.factor(cluster))) +
  geom_violin(trim = FALSE, fill = "gray") +
  geom_boxplot(width = 0.05) +
  theme(panel.border = element_rect(colour = "black", fill = NA, linewidth = 0.5)) #+  
     #coord_cartesian(ylim = c(0, 100))

```

#Test various methods to assess optimal number of clusters
#Silhouette Method 
```{r}

library(factoextra)

fviz_nbclust(master_random_numbers, kmeans, method = "silhouette") + 
  ggtitle("Silhouette Method for Optimal k") 



```

#elbow method
```{r}
library(cluster)
library(factoextra)

#create plot of number of clusters vs total within sum of squares
fviz_nbclust(master_random_numbers, kmeans, method = "wss") + 
  #geom_vline(xintercept = 2, linetype = 2 ) + 
  labs(subtitle = "Elbow method")
```


#gap stat method 
```{r}

set.seed(24)
gap_stat <- cluster::clusGap(master_random_numbers, FUN = kmeans, nstart = 25, K.max = 10, B = 50)

factoextra::fviz_gap_stat(gap_stat)
```
# Here were randomly generating 2 columns of rand numbers 
```{r}
height <- c(170,120,10,300,375,200,42,500)
measures <- as.data.frame(height)
measures$width <- c(74,24,1,180, 207, 120, 10, 357)


#set random parameters so theyre the same for each run
sd_value <- 1
num_samples <- 200
master_random_numbers <- data.frame()
#master_random_numbers <- c()
for(k in 1:length(measures$height)){ #k = 1
  loop.mean_height <- measures$height[k]
  #could do loop.mean_height <- measures{k, 1}
  loop.mean_width <- measures$width[k]
  #set seed
  set.seed(11)
random_numbers_height <- rnorm(num_samples, mean = loop.mean_height, sd = sd_value)
random_numbers_width <-pmax(rnorm(num_samples, mean = loop.mean_width, sd = sd_value),0)

random_numbers_comb <- cbind.data.frame(random_numbers_height, random_numbers_width )

master_random_numbers <- rbind.data.frame(master_random_numbers, random_numbers_comb)
  #master_random_numbers <- c(master_random_numbers, random_numbers )
}

head(master_random_numbers)

```








