---
title: "week5 knit file"
author: "Xavier Francis"
date: "2025-02-17"
output: html_document
---



#Load packages & data
```{r}
library(tidyverse)
library(factoextra)
library(cluster)
library(ggdendro)

```

# LOOP: ONE MEASURE
We want to write a loop to generate random data with different numbers of groups and different parameters and save the random numbers

```{r}
KEY.means <- c(42, 156, 675, 400, 3232, 7000)

#set the other random parameters, these are the same for each iteration
sd_value <- 1
num_samples <- 200

#write the data
random.numbers <- data.frame() #running this clears your dataframe

#LOOP
for(KEY in 1:length(KEY.means)){ #KEY=1
  loop.mean <- KEY.means[KEY]
  #set the seed
  set.seed(42)
  
  #radomize data
  random <- rnorm(num_samples, mean = loop.mean, sd = sd_value)
  random <- as.data.frame(random)

  #save the data
  random.numbers <- rbind.data.frame(random.numbers, random)
}

head(random.numbers )
```


## Elbow method

```{r}
test<- factoextra::fviz_nbclust(random.numbers, kmeans, method = "wss") +
  labs(subtitle = "Elbow method")

test #print the plot

names(test) #this shows you the parts contained in the output
test$data #this shows you the y axis values for each of the clusters in your output
```

## Silhouette method
```{r}
#library(factoextra)
factoextra::fviz_nbclust(random.numbers, kmeans, method = "silhouette")+
  labs(subtitle = "Silhouette method")
```


## Gap statistic
```{r}
#library(cluster)
# compute gap statistic
set.seed(123)
gap_stat <- cluster::clusGap(random.numbers, FUN = kmeans, nstart = 25,
                    K.max = 10, B = 50)
factoextra::fviz_gap_stat(gap_stat)

```



# LOOP: TWO MEASURES
We want to write a loop to generate random data with different numbers of groups and different parameters and save the random numbers

```{r}
height <- c(42, 156, 675, 400, 3232, 7000)

KEY.measures <- as.data.frame(height)

set.seed(42)
KEY.measures$width <- KEY.measures$height/(rnorm(1, mean=KEY.measures$height/2, sd=12))

KEY.measures
#set the other random parameters, these are the same for each iteration
sd_value <- 1
num_samples <- 200

#save the data
random.measures <- data.frame() #running this clears your dataframe

#LOOP
for(KEY in 1:nrow(KEY.measures)){ #KEY=1
  loop.mean1 <- KEY.measures[KEY, 1]
  loop.mean2 <- KEY.measures[KEY, 2]

  #radomize data
  set.seed(42)
  random.height <- rnorm(num_samples, mean = loop.mean1, sd = sd_value)
  random.height <- as.data.frame(random.height)
  
  set.seed(42)
  random.width <- rnorm(num_samples, mean = loop.mean2, sd = sd_value)
  random.width <- as.data.frame(random.width)

  #bind two measures together
  loop.randoms <- cbind.data.frame(random.height, random.width)
  #save the data
  random.measures <- rbind.data.frame(random.measures, loop.randoms)
}


glimpse(random.measures) #view head of output  
```
#kmeans with 2 random measures data
```{r}
kmeans(random.measures, 6)
```
#how do we partition the data?
 want to split data into 2 equal halves (random split)
```{r}
#number of rows
half_row <- nrow(random.measures)/2
full_row <- nrow(random.measures)

rand_measures1 <- random.measures[1:half_row,]

rand_measures2 <- random.measures[(half_row +1):full_row,]

#could also do
random.measures$split_ID <- sample(c(rep(1,600), rep(2,600)))
head(random.measures)

part_1 <- subset(random.measures, split_ID == 1)
part_2 <- subset(random.measures, split_ID == 2)

head(part_1)
head(part_2)

unique(part_1$split_ID)
```
