---
title: "ADA Week 8 Mantel"
author: "Xavier Francis"
date: "2025-03-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# What does a Mantel test do?

https://uw.pressbooks.pub/appliedmultivariatestatistics/chapter/mantel-test/: "The Mantel test examines the correlation between two distance matrices from the same samples.  The distance matrices can be based on actual data (species abundances, environmental measurements, spatial coordinates) or hypothesized data (e.g., dummy variables coding for treatments).  For example, Mantel tests are commonly used to test for spatial autocorrelation..."


# What is a Mantel test testing for? 
https://jkzorz.github.io/2019/07/08/mantel-test.html

https://besjournals.onlinelibrary.wiley.com/doi/full/10.1111/2041-210X.12425: "...Mantel test, which is a test of the absence of relationship between the dissimilarities in two dissimilarity matrices. A correct formulation of H0 for the Mantel test is the following: ‘H0: The distances among objects in matrix DY are not (linearly or monotonically) related to the corresponding distances in DX’ (Legendre & Legendre 2012, p. 600; italics added for emphasis). Similar formulations of the Mantel null hypothesis are found in Legendre (2000, p. 41): ‘The simple Mantel test is a procedure to test the hypothesis that the distances among objects in a [distance] matrix A are linearly independent of the distances among the same objects in another [distance] matrix B’ and in Legendre & Fortin (2010, p. 835)."


# How are Mantel results interpreted?
https://uw.pressbooks.pub/appliedmultivariatestatistics/chapter/mantel-test/
A significant Mantel test will tell you that the distances between samples in one matrix are correlated with the distances between samples in the other matrix. Therefore, as the distance between samples increases with respect to one matrix, the distances between the same samples also increases in the other matrix. 

The test statistic ranges from -1 (the two sets of distances are negatively correlated) to +1 (the two sets of distances are positively correlated).


# What are the assumptions for Mantel tests?
https://uw.pressbooks.pub/appliedmultivariatestatistics/chapter/mantel-test/
Mantel tests assume that the matrices are independent and that the correlations are linear.  Of course, since the matrices are being compared to one another, it is also essential that samples are listed in the same order in both matrices.


# Examples of uses:

https://uw.pressbooks.pub/appliedmultivariatestatistics/chapter/mantel-test/: "Mantel tests are generally used in two ways in community ecology: To compare two independent empirical matrices.  For example, spatial autocorrelation can be explored by comparing a distance matrix based on some response with a distance matrix based on the Euclidean distance between the sample units as determined from their spatial coordinates. To assess the goodness of fit between data and an a priori model (i.e., one not generated by this dataset).  This is how a Mantel test is used to compare a priori groups.  In this case, the second distance matrix is based on hypothesized data about the sample units (e.g., dummy variables coding for treatments)."

# Basic Procedure: (from https://uw.pressbooks.pub/appliedmultivariatestatistics/chapter/mantel-test/)

1. Convert each data matrix to a dissimilarity matrix.  Use an appropriate distance measure for each matrix; the measures do not have to be the same.

2. Calculate the correlation between the two dissimilarity matrices.  This is the test statistic.  As a correlation, this always ranges between -1 and +1.

3. Assess statistical significance via a permutation test:

a) Shuffle the row and column identities of one dissimilarity matrix, permuting row and column identities together so that matrix symmetry is preserved.  For example, if row A is moved down four positions, column A is also moved to the right four positions.  Equivalently (and easier to describe), permute the rows of one of the original data matrices and then recalculate the dissimilarity matrix.

b) Recalculate the correlation between the two dissimilarity matrices.  Save this value.

c) Re-shuffle and recalculate the specified number of times.  The permutations produce a sampling distribution of correlation values against which the actual test statistic is compared.

d) Calculate the P-value as the proportion of permutations that yielded equal or stronger correlations than the actual data did.  Note that Mantel tests generally result in a one-tailed test, as in this case.  A two-tailed test would be used if we had no a priori suspicion about whether the correlation would be positive or negative.  To obtain the P-value for a two-tailed test, we would calculate the proportion of permutations whose correlation coefficient (in absolute value) was greater than the test statistic.

https://rdrr.io/cran/ecodist/man/mantel.html

# Load packages
```{r}
library(tidyverse)
library(vegan)
```
# How to make distance matrices
https://www.geeksforgeeks.org/how-to-create-a-distance-matrix-in-r/
```{r}
# Create a matrix of data 
data <- matrix(c(1, 2, 3, 4, 5, 6, 7, 8), 
               nrow = 4, ncol = 2) 

#plot the data
data.d <- as.data.frame(data)
ggplot(data.d, aes(x=V1, y=V2)) +
  geom_point() +
  lims(x=c(0,8), y=c(0,8))+
  theme(aspect.ratio = 1,
        panel.border = element_rect(colour = "black", fill=NA, linewidth=0.5))

```


## Calculate distance
```{r}
# Create an Euclidean distance matrix 
euclidean_matrix <- dist(data) 
  
# Print the distance matrix 
print(euclidean_matrix) 
```
Euclidean distance is a measure of the straight-line distance between two points in Euclidean space. This output is a distance matrix where the value at i, j is the Euclidean distance between rows i and j of the data matrix. The dist() function calculates the Euclidean distance by default, so you don’t need to specify the method as “euclidean” explicitly.


```{r}
data1 <- matrix(c(seq(from = 1, to = 42, length.out = 50),
                  seq(from = 4, to = 30, length.out = 50)), 
                  nrow = 50, ncol = 2) 

data2 <- matrix(c(seq(from = 1, to = 56, length.out = 50),
                  seq(from = 4, to = 100, length.out = 50)), 
                  nrow = 50, ncol = 2) 

```

```{r}

#plot the data
data.d1 <- as.data.frame(data1)
data.d2 <- as.data.frame(data2)

ggplot() +
  geom_point(data=data.d1, aes(V1,V2), color="red") +
  geom_point(data=data.d2, aes(V1,V2), color="blue") +
  #lims(x=c(0,8), y=c(0,8))+
  theme(aspect.ratio = 1,
        panel.border = element_rect(colour = "black", fill=NA, linewidth=0.5))


```
## Find two distances
```{r}
euclidean_matrix1 <- dist(data1, diag = TRUE) 
euclidean_matrix2 <- dist(data2, diag = TRUE) 


mantel.result <- vegan::mantel(euclidean_matrix1, euclidean_matrix2, method="pearson", permutations=1000)
names(mantel.result)

#extract r and p-values from results
mantel.result$statistic
result.label <- paste0("R=", mantel.result$statistic, ", p-value=", round(mantel.result$signif,4))
result.label
```

## Plot results
```{r}
perm.data <- as.data.frame(mantel.result$perm)
colnames(perm.data) <- "perms"

ggplot(perm.data, aes(x=perms)) + 
geom_histogram(color="black", fill="lightgrey")+
  geom_vline(xintercept=mean(perm.data$perms), color="black", lty=2) +
  geom_vline(xintercept=mantel.result$statistic, color="red")+
  labs(title=paste0("Mantel test results: ", result.label)) +
  theme(panel.border = element_rect(colour = "black", fill=NA, linewidth=0.5))
```

Broken mantel examples
```{r}

# Set the seed for reproducibility
set.seed(62)
 
# Generate  random numbers from a normal distribution
mean_value <- 35
sd_value <- 96
num_samples <- 50
random_numbers1 <- rnorm(num_samples, mean = mean_value, sd = sd_value)

set.seed(624)
mean_value2 <- -90
sd_value2 <- 51
#num_samples <- 400
random_numbers2 <- rnorm(num_samples, mean = mean_value2, sd = sd_value2)

data1 <- data.frame(random_numbers1, random_numbers2)






# Set the seed for reproducibility
set.seed(5)
 
# Generate  random numbers from a normal distribution
mean_value <- 10
sd_value <- 9
num_samples <- 50
random_numbers3 <- rnorm(num_samples, mean = mean_value, sd = sd_value)

set.seed(9)
mean_value2 <- 20
sd_value2 <- 85
#num_samples <- 400
random_numbers4 <- rnorm(num_samples, mean = mean_value2, sd = sd_value2)



data2 <- data.frame(random_numbers3, random_numbers4)



#convert to matrix
matrix1 <- as.matrix(data1)
matrix2 <- as.matrix(data2)
```

```{r}

#plot the data
data.d1 <- as.data.frame(matrix1)
data.d2 <- as.data.frame(matrix2)

ggplot() +
  geom_point(data=data.d1, aes(random_numbers1,random_numbers2), color="red") +
  geom_point(data=data.d2, aes(random_numbers3, random_numbers4), color="blue") +
  #lims(x=c(0,8), y=c(0,8))+
  theme(aspect.ratio = 1,
        panel.border = element_rect(colour = "black", fill=NA, linewidth=0.5))


```
## Find two distances
```{r}
euclidean_matrix1 <- dist(data1, diag = TRUE) 
euclidean_matrix2 <- dist(data2, diag = TRUE) 


mantel.result <- vegan::mantel(euclidean_matrix1, euclidean_matrix2, method="pearson", permutations=1000)
names(mantel.result)

#extract r and p-values from results
mantel.result$statistic
result.label <- paste0("R=", mantel.result$statistic, ", p-value=", round(mantel.result$signif,4))
result.label
```

## Plot results
```{r}
perm.data <- as.data.frame(mantel.result$perm)
colnames(perm.data) <- "perms"

ggplot(perm.data, aes(x=perms)) + 
geom_histogram(color="black", fill="lightgrey")+
  geom_vline(xintercept=mean(perm.data$perms), color="black", lty=2) +
  geom_vline(xintercept=mantel.result$statistic, color="red")+
  labs(title=paste0("Mantel test results: ", result.label)) +
  theme(panel.border = element_rect(colour = "black", fill=NA, linewidth=0.5))
```

#on tails but not sigifcant. Still lies within what is expected at chance
