---
title: "ADA_week_6_day_2"
author: "Xavier Francis"
date: "2025-02-18"
output: html_document
---



#load packages
```{r}
library(tidyverse)
library(car)
#install.packages("ggbiplot")
library(ggbiplot)
library(factoextra)
library(ggplot2)
```

```{r}
whales_df <- read.csv("C:/Users/Xmanf/Desktop/Advanced data analysis stuff/whale_pca2.csv", header=TRUE, stringsAsFactors=FALSE)

```

```{r}

names(whales_df) <- c("species", "longth", "weight_tons", "blubber_thickness", "fin_size","dive_depth", "swim_speed", "lung_capac", "gest_period", "lifespan")


```


# Run PCA
We run the PCA with prcomp(). Because the variable scales are not all commensurable, it is important to scale them to equal variance." (scale=TRUE does this)
```{r}
whale.pca <- prcomp (~ . ,
                    data=whales_df[2:10],
                    na.action=na.omit,  # not actually necessary: we removed NA
                    scale. = TRUE)

whale.pca

```
## How much variation is explained by each dimension?
Visualize eigenvalues (scree plot). Show the percentage of variances explained by each principal component.
```{r}
factoextra::fviz_eig(whale.pca)
```


## Extract the data from the pca output
```{r}
pca.summ <- summary(whale.pca)
pca.summ
names(pca.summ)
pca.summ$importance
pca.summ$importance[2,] #extract just the row with the percent explained by each dimension
```

## Plot as biplot
```{r}
ggbiplot::ggbiplot(whale.pca, obs.scale = 1, var.scale = 1,
         groups = whales_df$species, 
         ellipse = TRUE, circle = TRUE) +
  scale_color_discrete(name = 'Whale Species') +
  theme_minimal() +
  theme(legend.direction = 'horizontal', legend.position = 'top') 
```


## Contribution of each variable  https://www.datacamp.com/tutorial/pca-analysis-r
"The goal of the third visualization is to determine how much each variable is represented in a given component. Such a quality of representation is called the Cos2 and corresponds to the square cosine, and it is computed using the fviz_cos2 function. A low value means that the variable is not perfectly represented by that component. A high value, on the other hand, means a good representation of the variable on that component."
```{r}
factoextra::fviz_cos2(whale.pca, choice = "var", axes = 1)
factoextra::fviz_cos2(whale.pca, choice = "var", axes = 2)
#fviz_cos2(whale.pca, choice = "var", axes = 1:2)

```

#testing corelation 
```{r}


cor.test(whales_df$lifespan, whales_df$gest_period )
cor.test(whales_df$swim_speed, whales_df$fin_size )



cor.test(whales_df$dive_depth, whales_df$blubber_thickness )

```



